---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
[![Travis-CI Build Status](https://travis-ci.org/jonesor/Rage.svg?branch=master)](https://travis-ci.org/jonesor/Rage)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/jonesor/Rage?branch=master&svg=true)](https://ci.appveyor.com/project/jonesor/Rage)


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
options(digits = 4)
```

# Rage

An R package for matrix population model analysis. Note this package is at an early stage of development, and may contain bugs.


## Installation

Install from GitHub with:

```{r, eval=FALSE}
devtools::install_github("jonesor/Rage")
#
# or
#
install.packages("remotes") # smaller and quicker to install than devtools
remotes::install_github("jonesor/Rage")
```

## Usage

```{r}
library(Rage)
```

First we need a matrix population model separated into the growth/survival component and sexual reproduction component (possibly also a clonal component, but we'll skip that here).

```{r}
matU <- rbind(c(0.1,   0,   0,   0), # growth/survival component
              c(0.6, 0.2, 0.1,   0),
              c(  0, 0.5, 0.5, 0.1),
              c(  0,   0, 0.3, 0.8))

matF <- rbind(c(  0,   0, 0.2, 0.6), # sexual reproduction component
              c(  0,   0, 0.1, 0.2),
              c(  0,   0,   0,   0),
              c(  0,   0,   0,   0))
```

Now we can get to work. Here are just a few example analyses.
<br>

Calculate life expectancy:

```{r}
lifeExpectancy(matU)
```

Calculate net reproductive rate:

```{r}
R0(matU, matF)
```

Produce a life table:

```{r}
makeLifeTable(matU, matF, nSteps = 15)
```

Collapse the matrix population model to two stages, pre-reproductive and reproductive, and conduct perturbation analyses on the vital rates of the collapsed matrix:

```{r}
# pre-reproductive stages are 1 and 2, reproductive stages are 3 and 4
mpm_collapse <- collapseMatrix(matU, matF, collapse = list(1:2, 3:4))

# perturbation analysis (S- represent sensitivities, and E- elasticities)
vitalRatePerturbation(mpm_collapse$matU, mpm_collapse$matF)
```

