---
title: "Getting started with Rage"
author: "Patrick Barks"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with Rage}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(digits = 4)
#The following line seems to be required by pkgdown::build_site(), but causes build to break with R-CMD-CHECK on GH
#knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

`Rage` provides tools for manipulating and analysing matrix population models (MPMs). This vignette introduces the structure of the input and highlights key analysis functions across the package.

# Loading MPMs and basic anatomy

We'll begin by loading one of the example MPMs called `mpm1` included in the `Rage` package and which can be retrieved using the base 'data' function.

```{r}
library(Rage)  # load Rage
data(mpm1)     # load data object 'mpm1'
mpm1           # display the contents
```

The object `mpm1` is a list containing square matrices as elements that are the decomposition products of an MPM projection matrix. `Rage` functions use one or more of these components in analyses and recognizes the biological meaning of each list element by its name:

- `matU` is the __U__ matrix, which is the __growth/survival__ component of an MPM 
containing transitions related to progression, stasis and retrogression. This is how individuals move among ages or life stages.
- `matR` is the __R__ matrix, which is the __reproductive__ component of an MPM,
containing transitions due to reproduction; either sexual, clonal, or both.
When the mode of reproduction is known (e.g., a species only reproduces sexually or the sexual and clonal offspring have been counted separately), `matR` can be substituted with:
- `matF` is the __F__ matrix, which is the __sexual reproduction__ or fecundity 
component of the MPM containing transitions due to sexual reproduction. This is one way that new individuals are produced.
- `matC` is the __C__ matrix, which is the __clonal__ component of an MPM 
containing transitions due to clonal reproduction.

For any life history, we can reconstruct the full projection matrix, __A__, by adding together the components. For `mpm1`,  __A__ = __U__ + __F__. A population with separately counted sexual and clonal offspring production would be __A__ = __U__ + __F__ + __C__.

`Rage` functions currently accept MPM components as arguments; therefore, it is not necessary to group the `mat*`s together in a list. We anticipate that future package releases will define methods that accept objects of classes `matrix`, `CompadreMat`, and `CompadreDB` for compatibility with the [COM(P)ADRE databases](https://compadre-db.org/) and its accessor package [`Rcompadre`](https://github.com/jonesor/Rcompadre).

The remaining numbered sections introduce the key functions of `Rage` across five broad categories:

1. [deriving life history traits](#lifehisttraits)
2. [deriving life tables or life table components](#lifetablecomponents)
3. [deriving vital rates](#vitalrates)
4. [perturbation analyses](#perturbanalyses)
5. [manipulating or transforming MPMs](#maniptransform)

# 1. Deriving life history traits {#lifehisttraits}

What is the life expectancy of an individual? At what age will it begin to reproduce? How likely is it to survive to reproduction? What is the generation time? These high level questions address the population-level life history traits that emerge from aggregating individual-level demographic rates, and tracing trajectories through the life cycle.

Life history traits that do not involve reproduction require only a __U__ matrix (supplied to the function's `matU` argument) and the stage class that represents the starting point for the calculation (supplied to the `start` argument). In the example MPM, a natural starting point for calculating life expectancy could be from either the seed stage or post-recruitment. Both calculations are possible by modifying the `start` argument. Longevity, defined as the age at which survivorship falls below a user-defined threshold, follows a similar pattern.
```{r}
life_expect(matU = mpm1$matU, start = 1)  # life expectancy from "seed" stage
life_expect(matU = mpm1$matU, start = 2)  # life expectancy from "small" stage
longevity(matU = mpm1$matU, start = 2, lx_crit = 0.05)  # post-germination years until survivorship falls below 5%
```

An expanded set of life history traits can be calculated by incorporating a reproduction matrix to the `matR` argument. In our example, we'll assume that reproduction only happens sexually and therefore we can substitute __F__ in place of __R__. For species with both modes of reproduction, using __F__ or __C__ alone in these functions allows for a more nuanced assessment of reproductive life history. For reproductive traits that aggregate across the life cycle, we need not supply a `start` stage.

```{r}
mature_age(matU = mpm1$matU, matR = mpm1$matF, start = 2)     # post-germination years to first reproduction
mature_prob(matU = mpm1$matU, matR = mpm1$matF, start = 2)    # post-germination Pr(survival to first repro)
net_repro_rate(matU = mpm1$matU, matR = mpm1$matF)            # net reproductive rate (aggregate)
gen_time(matU = mpm1$matU, matR = mpm1$matF)                  # generation time (aggregate)
```

Other life history traits are calculated from a life table rather than an MPM,
in which case we can first use the `mpm_to_` group of functions to derive the
necessary life table components: (st)age-specific survivorship (*lx*), survival probability (*px*), mortality hazard (*hx*), and reproduction (*mx*).
```{r}
lx <- mpm_to_lx(matU = mpm1$matU, start = 2)
px <- mpm_to_px(matU = mpm1$matU, start = 2)
hx <- mpm_to_hx(matU = mpm1$matU, start = 2)
mx <- mpm_to_mx(matU = mpm1$matU, matR = mpm1$matF, start = 2)

# then calculate life history traits
entropy_k(lx)       # Keyfitz' entropy
entropy_d(lx, mx)   # Demetrius' entropy
shape_surv(lx)      # shape of survival/mortality trajectory
shape_rep(mx)       # shape of fecundity trajectory
```

# 2. Life tables and the quasi-stationary distribution {#lifetablecomponents}

Some MPMs are parameterized with a stasis loop at the maximum stage class, which
can lead to apparent plateaus in mortality or fertility trajectories derived
using age-from-stage methods. The function `qsd_converge()` can be used to
identify the time it takes for a cohort to reach the quasi-stationary
distribution (QSD). This quantity can then be used to subset age trajectories of
mortality or fertility to periods earlier than the QSD, so as to avoid
artefactual plateaus in mortality or fertility.

```{r, warning=FALSE, message=FALSE}
# derive life table from MPM
lt <- mpm_to_table(mpm1$matU, start = 2)

# calculate time to QSD
(q <- qsd_converge(mpm1$matU, start = 2))

# plot mortality trajectory w/ vertical line at time to QSD
par(mar = c(4.5, 4.5, 1, 1))
plot(qx ~ x, data = lt, type = "l", ylim = c(0, 0.65))
abline(v = q, lty = 2)
```

From the life table derived from `mpm1`, we can see a plateau in the mortality
rate (qx) beginning around age 5. However, this plateau corresponds to the QSD
and is therefore probably an artefact of the stasis loop rather than a
biological reality for the population represented by `mpm1`.

One approach to accounting for this artefactual plateau in subsequent life
history calculations is to limit our life table to the period prior to the QSD.

```{r}
# calculate the shape of the survival/mortality trajectory
shape_surv(lt$lx)       # based on full lx trajectory
shape_surv(lt$lx[1:q])  # based on lx trajectory prior to the QSD
```

# 3. Standardized vital rates {#vitalrates}

The transition rates that make up MPMs generally reflect products of two or more
vital rates (sometimes called ‘lower-level vital rates’). Assuming a
post-breeding census design, we can retroactively break apart each transition
rate into at least two vital rate components: survival, and ‘something’
conditional on survival. That ‘something’ might be growth, shrinkage, stasis,
dormancy, fecundity, or clonality.

## Stage-specific vital rates (vector)

To summarize vital rates _within_ stage classes, we can use the `vr_vec_` group
of functions. We'll use the `exclude` argument here to exclude certain stage
classes ('seed' and 'dormant') from the calculation of certain vital rates (e.g.
we don't consider the large-to-dormant transition to actually represent
'growth').

```{r}
vr_vec_survival(mpm1$matU)
vr_vec_growth(mpm1$matU, exclude = c(1, 5))
vr_vec_shrinkage(mpm1$matU, exclude = 5)
vr_vec_stasis(mpm1$matU)
vr_vec_dorm_enter(mpm1$matU, dorm_stages = 5)
vr_vec_dorm_exit(mpm1$matU, dorm_stages = 5)
vr_vec_reproduction(mpm1$matU, mpm1$matF)
```

## MPM-specific vital rates (scalar)

To summarize vital rates _across_ stage classes, we can use the `vr_` group of
functions. By default these functions take a simple average of the
stage-specific vital rates produced by the corresponding `vr_vec_` function.
However, here we'll demonstrate how to specify a _weighted_ average across
stages, based on the stable stage distribution at equilibrium (_w_).

```{r}
# derive full MPM (matA)
mpm1$matA <- mpm1$matU + mpm1$matF

# calculate stable stage distribution at equilibrium using popbio::stable.stage
library(popbio)
w <- popbio::stable.stage(mpm1$matA)

# calculate MPM-specific vital rates
vr_survival(mpm1$matU, exclude_col = c(1, 5), weights_col = w)
vr_growth(mpm1$matU, exclude = c(1, 5), weights_col = w)
vr_shrinkage(mpm1$matU, exclude = c(1, 5), weights_col = w)
vr_stasis(mpm1$matU, exclude = c(1, 5), weights_col = w)
vr_dorm_enter(mpm1$matU, dorm_stages = 5, weights_col = w)
vr_dorm_exit(mpm1$matU, dorm_stages = 5, weights_col = w)
vr_fecundity(mpm1$matU, mpm1$matF, weights_col = w)
```

Note how we've chosen to exclude the 'seed' and 'dormant' stage classes from our
vital rate summaries, because we consider these to be special classes (e.g.
'growth' from the 'seed' stage is really 'germination', which we may think of as
separate from somatic growth from 'small' to 'medium', or 'medium' to 'large').

# 4. Perturbation analyses {#perturbanalyses}

The `perturb_matrix()` function measures the response of a demographic statistic
to perturbation of individual matrix elements (i.e. sensitivities and
elasticities). The `perturb_vr()` and `perturb_trans()` functions implement
perturbation analyses by vital rate type (survival, growth, etc.) and transition
type (stasis, retrogression, etc.), respectively.

```{r}
# matrix element perturbation
perturb_matrix(mpm1$matA, type = "sensitivity")

# vital rate perturbation
# (we use as.data.frame here for prettier printing)
as.data.frame(perturb_vr(mpm1$matU, mpm1$matF, type = "sensitivity"))

# transition type perturbation
as.data.frame(perturb_trans(mpm1$matU, mpm1$matF, type = "sensitivity"))
```

# 5. Transforming MPMs {#maniptransform}

`Rage` includes a variety of functions that can be used to manipulate or transform
MPMs. For example, we can collapse an MPM to a smaller number of stage classes
using `mpm_collapse()`.

```{r}
# collapse 'small', 'medium', and 'large' stages into single stage class
col1 <- mpm_collapse(mpm1$matU, mpm1$matF, collapse = list(1, 2:4, 5))
col1$matA
```

The transition rates in the collapsed matrix are a weighted average of the
transition rates from the relevant stages of the original matrix, weighted by
the stable distribution at equilibrium. This process guarantees that the
collapsed MPM will retain the same population growth rate as the original.
However, other demographic and life history characteristics will not necessarily
be preserved.

```{r}
# compare population growth rate of original and collapsed MPM (preserved)
popbio::lambda(mpm1$matA)
popbio::lambda(col1$matA)

# compare net reproductive rate of original and collapsed MPM (not preserved)
net_repro_rate(mpm1$matU, mpm1$matF)
net_repro_rate(col1$matU, col1$matF)
```

# Appendix: Complete list of functions

| Category | Function | Description   |
|:---------------|:-------------|:-------------------------------------|
| Life history traits  | `life_expect`    | Life expectancy |
|                      | `longevity`      | Longevity |
|                      | `net_repro_rate` | Net reproductive rate |
|                      | `gen_time`       | Generation time |
|                      | `mature_age`     | Age at reproductive maturity |
|                      | `mature_prob`    | Probability of reaching reproductive maturity |
|                      | `mature_distrib` | Stage distribution of reproductive maturity |
|                      | `entropy_d`      | Demetrius' entropy |
|                      | `entropy_k`      | Keyfitz' entropy |
|                      | `shape_surv`     | Shape of survival/mortality trajectory |
|                      | `shape_rep`      | Shape of fecundity trajectory |
| Life table           | `mpm_to_table`   | MPM to life table |
|                      | `mpm_to_lx`      | MPM to survivorship trajectory |
|                      | `mpm_to_px`      | MPM to survival trajectory |
|                      | `mpm_to_hx`      | MPM to mortality hazard trajectory |
|                      | `mpm_to_mx`      | MPM to fecundity trajectory |
|                      | `lx_to_[px/hx]`  | Convert from survivorship trajectory |
|                      | `px_to_[lx/hx]`  | Convert from survival trajectory |
|                      | `hx_to_[lx/px]`  | Convert from mortality hazard trajectory |
|                      | `qsd_converge`   | Time to quasi-stationary distribution |
| Vital rates          | `vr_[...]`       | MPM-averaged vital rates
|                      | `vr_vec_[...]`   | Stage-averaged vital rates
|                      | `vr_mat_[...]`   | Survival-independent vital rates |
| Perturbation         | `perturb_matrix`     | Perturbation analysis of whole matrix |
|                      | `perturb_trans`      | Perturbation analysis of transition types |
|                      | `perturb_vitals`     | Perturbation analysis of vital rate types |
|                      | `perturb_stochastic` | Stochastic perturbation analysis |
| MPM transformation   | `mpm_split`          | Split MPM into survival and reproductive components |
|                      | `mpm_rearrange`      | Rearrange MPM to segregate reproductive stages |
|                      | `mpm_collapse`       | Collapse MPM to smaller number of stages |
|                      | `mpm_standardize`    | Collapse MPM to standardized set of stages |
|                      | `standard_stages`    | Group stages into standardized sets |
|                      | `repro_stages`       | Identify reproductive stages |
|                      | `plot_life_cycle`    | Plot a life cycle diagram |
