---
title: "Ternary Plots"
author: "Rob Salguero-Gomez and Owen Jones"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ternary Plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here we produce a ternary plot *a la* Silvertown & Franco (1993) with population growth rate as the "fourth" dimension. We will use functions from the `popdemo`, `Rage` and `ggtern` packages, so first we need to load those packages:

```{r, message=FALSE}
library(Rcompadre)
library(popdemo)
library(Rage)
library(ggtern)
```

Next, we will load the dataset directly from the COMPADRE online repository. Then, we will subset making sure that we only select matrix population models (*MPMs*, hereafter) fulfilling two conditions. The first one is that the components of the MPM **A** (e.g. **A** = **U** + **F** + **C**; **U**: survival-dependent matrix; **F**: sexual reproduction matrix; **C**: clonal reproduction matrix) have already been split so we can automatically calculate vital-rate dependent processes (e.g. elasticities of population growth rate (*$\lambda*). The second condition is that the MPM **A** must be non-ergodic (See Caswell 2001 for further details), so there is a unique solution for the dominant eigenvalue of **A**.

```{r load example COMPADRE data, eval=TRUE}
data(Compadre)
Compadre <- cdb_flag(Compadre)
Compadre <- subset(Compadre, MatrixSplit == "Divided" & check_ergodic == TRUE)
```

We can calculate elasticities of population growth rate to changes in the matrix elements using the `elas` function from the `popdemo` package. For example, the elasticity matrix for the first **A** matrix in this example database looks like this:

```{r elas, eval=TRUE}
popdemo::elas(matA(Compadre)[[1]])
```

Because the subset of MPMs here has split **A** into the matrices **U**, **F**, and **C**, we can now also automatise the calculation of elasticities, regardless of the varying degree of life cycle complexities represented in these MPMs. Therefore, we can classify the elements of the matrix to changes in stages (which we can further identify as stasis, progression, and retrogression), sexual reproduction, and clonal reproduction. 

We can use the `matrixElementPerturbation` function from the `Rage` package to conduct this element-by-element elasticity analysis. The function outputs both sensitivities (prefixed with S) and elasticities (prefixed with E) as follows:


```{r}
mat_a <- matA(Compadre)[[1]]

perturb_matrix(mat_a, type = "elasticity")

#Rage::matrixElementPerturbation(matU = matU(Compadre)[[1]],    This function used to be able to accommodate partioning
#                                matF = matF(Compadre)[[1]],    into U F and C so we coudl partition this... but it's now gone
#                                matC = matC(Compadre)[[1]])    in perturb_matrix.... why?
```

We can use a `for` loop to run through each set of matrices in turn to calculate the summed elasticity for survival (S), growth (G) and reproduction (R) like this:

```{r}
Amats <- matA(Compadre)
Umats <- matU(Compadre)
Fmats <- matF(Compadre)
Cmats <- matC(Compadre)

output <- data.frame(S=rep(NA,length(Umats)),G=NA,R=NA,lam=NA)

for(i in 1:length(Umats)){
  temp <- perturb_vr(Umats[[i]],   #The same issue as above
                     Fmats[[i]],
                     Cmats[[i]],
                     type = "elasticity")
  
  output$S[i] <- temp$survival
  output$G[i] <- temp$growth + temp$shrinkage
  output$R[i] <- temp$fecundity + temp$clonality
  
  #Calculate growth rate
  output$lam[i] <- popdemo::eigs(Amats[[i]], "lambda")
}
```

Let's take a look at this output:

```{r}
head(output)
```

Now we have elasticities for our three demographic processes we can place these onto a ternary plot. However, we should first scale the elasticities so that they sum to 1 - this is necessary because of possible rounding issues:

```{r}
output[,1:3] <- t(apply(output[,1:3], 1, function(x) x/sum(x)))
```

Now for the plot:

```{r}
B<-ggtern::ggtern(data = output,
                  aes(x = R,
                      y = G,
                      z = S,
                      colour = lam))  +  
  geom_point() +    
  scale_color_viridis_c()+
  theme_showarrows() +
  theme_clockwise() 

B
```

Now you can try this, try using another variable, such as reactivity, or life expectancy as the "fourth" dimension instead of lambda. 

IN THE TEXT BELOW I RE-DO THE FUNCTIONALITY ABOVE AT THE VITAL RATE LEVEL. WE SHOULD FIRST FIX THE FUNCTION ABOVE SO IT DOES THE PARTITIONING OF MATRICES INTO SUBCOMPONENTS U F AND C

We can use the `perturb_vr` function from the `Rage` package to conduct the perturbations at the level of underlying vital rates rather than matrix elements. It is worth noting that in a MPM, most matrix elements are composites of vital rates. For instance, the element a[2,1] (second row, first column in the matrix **A**) is composed by the survival probability of the first stage *times* the probability that individuals in the first stage transition to the second one in the life cycle of the species. To proceed with the calculation of the vital rate perturbations:


```{r}

data(mpm1)

perturb_vr(mpm1$matU,
           mpm1$matF,
           mpm1$matF,
           type = "elasticity")
```

We can use a `for` loop to run through each set of matrices in turn to calculate the summed elasticity for (i) survival ($\sigma), (ii) growth ($\gamma) and shrinkage ($\rho), and (iii) sexual ($\phi) and clonal reproduction ($\kappa) like this:

```{r}
Amats <- matA(Compadre)
Umats <- matU(Compadre)
Fmats <- matF(Compadre)
Cmats <- matC(Compadre)

output <- data.frame(survival=rep(NA,length(Umats)),growthShrinkage=NA,reproduction=NA,lam=NA)
for(i in 1:length(Umats)){
  temp <- perturb_vr(Umats[[i]],
                     Fmats[[i]],
                     Cmats[[i]],
                     type = "elasticity")
  
output$survival[i] <- temp$survival
output$growthShrinkage[i] <- temp$growth + temp$shrinkage
output$reproduction[i] <- temp$fecundity + temp$clonality

#Calculate growth rate
output$lam[i] <- popdemo::eigs(Amats[[i]], "lambda")
}
```

Let's take a look at this output:

```{r}
head(output)
```

Now we have elasticities for our three collapsed vital rate processes we can place these onto a ternary plot. However, we should first scale the elasticities so that they sum to 1 - this is necessary because of possible rounding issues:

```{r}
output[,1:3] <- t(apply(output[,1:3], 1, function(x) x/sum(x)))
```

Now for the plot:

```{r}

B<-ggtern::ggtern(data = output,
                  aes(x = reproduction,
                      y = growthShrinkage,
                      z = survival,
                      colour = lam))  +  
  geom_point() +  
  scale_color_viridis_c()+
  theme_showarrows() +
  theme_clockwise() 

B
```
